---
title: "network-analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{network-analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tidyverse)
library(tidytacos)

urt.nf <- urt %>% 
filter_samples(location == "NF") %>%
add_taxon_name() 

```

This is a brief guide on how to perform network analysis on microbiome
count data using a similar strategy as performed in the [Isala study](https://www.nature.com/articles/s41564-023-01500-0) using tidytacos. 

The method clusters taxa based on their co-occurrence patterns across the samples. 

As per usual the test dataset used is a human microbiome samples from the upper respiratory
tract (URT), taken from [this
paper](https://www.frontiersin.org/articles/10.3389/fmicb.2017.02372/full)
by De Boeck et al. It contains nose as well as nasopharynx samples. Most
samples were taken using a swab method, but a minority was taking with
the aspirate method.

## Step 1: Creating a network

The first step in the analysis is creating a sparcc network from our taxonomic counts.
For this the [SpiecEasi package](https://github.com/zdk123/SpiecEasi) is used.

```{r, install-spieceasi}
devtools::install_github('zdk123/SpiecEasi')
```

```{r, sparcc-network }
urt.net <- urt.nf %>%
  network(
    rarefact = 0.05, # rarefaction level
    iter=20, # number of iterations of the outer loop of sparcc 
    inner_iter=10, # number of iterations of the inner loop of sparcc
    th=0.1, # correlations beneath this threshold are set to 0
    taxon_name=taxon_name)

# The network object is a list with two elements:
urt.net$Cor %>% head(2)# Correlation matrix
# urt.net$Cov (covariance matrix)
# urt.net$names (names of the taxa on x and y-axes)
```

# Step 2: Filtering the network matrix

Next up we filter the output correlation matrix to remove weak correlations.
For this we use the `filter_network` function, which has the `Matrix` package as a dependency.
We also install the `igraph` package, which can be used for the visualization of the network.
```{r, eval=FALSE}
install.packages("Matrix")
install.packages("igraph")
```

We remove weak correlations (here set to 0.1) take the absolute value of the correlation matrix and remove the diagonal.
This is all handled with the `filter_network` function, but we show the code here for clarity and to plot the network graph before clustering.
```{r, network-filter, fig.width=7, fig.height=7}
library(Matrix)
library(igraph)

urt.net.filt <- urt.net %>% filter_network(threshold = 0.1)
# this is what the function does:
se.sparcc.graph <- abs(urt.net$Cor) >= 0.1 # the treshold, absolutes as analysis is undirected
diag(se.sparcc.graph) <- 0 # remove diagonal correlations (= always 1)
sparcc.graph <- Matrix::Matrix(se.sparcc.graph, sparse = TRUE)
rownames(sparcc.graph) <- urt.net$names
graph <- igraph::graph_from_adjacency_matrix(sparcc.graph, mode = "undirected")
plot(graph)

```

# Step 3: Clustering of the network

Now the taxa in the network are clustered using an unsupervised clustering algorithm, the [Markov Cluster Algorithm (MCL)](https://micans.org/mcl/).
The `cluster_network` function has the `MCL` package as a dependency.

```{r, eval=FALSE}
install.packages("MCL")
```
```{r, network-clustering, fig.width=7, fig.height=7}
urt.net.clust <- urt.net.filt %>% 
  cluster_network(min_n=5, visualize=T)

urt.net.clust %>% 
  group_by(cluster) %>% 
  summarize(n=n())

```

Using the new clustered taxa ("eigentaxa") we can visualize the clusters in the microbiome profile.

```{r, fig.width=8, fig.height=5}
urt.nf$taxa <- urt.nf$taxa %>% 
  left_join(urt.net.clust, by=c("taxon_name"="taxon")) 
urt.nf <- urt.nf %>%
  add_taxon_name() %>% 
  mutate_taxa(cluster= ifelse(is.na(cluster), taxon_name, cluster)) %>% 
  set_rank_names(c("cluster")) %>% 
  aggregate_taxa(rank="cluster")
urt.nf  %>% tacoplot_stack()

urt.net.clust %>% filter(cluster == unique(urt.net.clust$cluster)[[4]])

```
Here cluster `r unique(urt.net.clust$cluster)[[4]]` and its composition are displayed.